What We Eval in the Shadows

### 2021-04-16 paper/main.tex
Update

In this paper, we provide a large-scale study of the use of the \eval function
in the R programming language. It is based primarily on a corpus of
\CranRunnableScripts scripts extracted from \CranPackages R packages. As
expected, \eval is used mainly in libraries and therefore we have found just
a few \eval call sites in scripts coming from Kaggle that we included to
represent code written by R practitioners. We used an instrumented version of
the R virtual machine, \rdyntrace, to capture all \eval calls traces including
the relevant side effects that happen within these calls. Several insights can
be learnt from the data.
%
First, \eval is widely used in the R core packages that come bundled with the
language. Because of that, there is hardly any R code that would not use it.
\eval is used to implement some basic functionality such as package loading.
Putting the core packages aside, in CRAN, \eval is used by \PkgPackagesRatio of
the available \CranPackages packages.
%
The code executed by R \eval has unusual expressive power. It can essentially
modify anything anywhere which is wary for both R developers and R compiler.
% TODO expressions passed to eval
Even more so in the case of side effects that can quite literally happen
anywhere. The data however suggest that this is not the case. The majority of
eval sites do not do any observable side effects and when they do, it happens
in a predictable environment.

>>>>

In this paper, we provide a large-scale study of the use of the \eval function
in the R programming language. It is based primarily on a corpus of
\CranRunnableScripts scripts extracted from \CranPackages R packages. As
expected, \eval is used mainly in libraries. We have found just a few \eval
call sites in scripts coming from large corpus of Kaggle that represents code
written by R practitioners. We used an instrumented version of the R virtual
machine, \rdyntrace, to capture all \eval calls traces including the relevant
side effects that happen within these calls. Several insights can be learnt
from the data.
%
First, \eval is widely used in the R core packages that come bundled with the
language. Because of that, there is hardly any R code that would not use it.
\eval is used to implement some basic functionality such as package loading.
Putting the core packages aside, in CRAN, \eval is used by \PkgPackagesRatio of
the available \CranPackages packages.
%
The code executed by R \eval has unusual expressive power. It can essentially
modify anything anywhere which is wary for both R developers and R compiler.
% TODO expressions passed to eval
Even more so in the case of side effects that can quite literally happen
anywhere. The data however suggest that this is not the case. The majority of
eval sites do not do any observable side effects and when they do, it happens
in a predictable environment.
